Parsing Postfix log files

### 2009-03-03 doc/logparser-architecture.tex
Dealt with some XXX items, and improved the pathological rules section.

XXX IS WEIGHT THE CORRECT WORD HERE\@?  SHOULD IT BE xxx WEIGHT\@?  GET
GREG TO PROOFREAD THIS SECTION\@.

A more complicated example is determining the weight of a compound given
its chemical formula.  A chemical formula is a sequence of one or two
character chemical element abbreviations, each optionally followed by a
number indicating how many of that element are present in the formula (if
omitted the number is assumed to be one).  Some parts of a formula may be
repeated: they are enclosed in parentheses, and the closing parenthesis is
followed by a number indicating how many are present in the formula.  A
monomer is a simple compound, i.e.\ it does not have any repeated portions;
compounds with repeated portions are called polymers.  E.g.\ glucose (a
monomer) has the chemical formula $C_{6}H_{12}O_{6}$ and hydrocarbon (a
polymer) is $CH_{3}{(CH_{2})}_{50}CH_{3}$.  The atomic weight of an
individual element is given in periodic tables, but one element in
particular will be important in the later explanation: Hydrogen,
abbreviated H, has an atomic weight of 1.  The weight of a given compound
is the sum of the atomic weights of each of the elements multiplied by the
number of times they occur; this is simple to calculate for a monomer, but
it is harder for a polymer because the weight of the repeated monomer must
be multiplied by the number of times the monomer is present in the compound
(e.g.\ for hydrocarbon the weight of $CH_{2}$ must be multiplied by $50$).

Calculating the weight of a monomer requires a single rule and action: the
action separates each element in the monomer, looks up its weight,
multiplies it by the number of times it is present (one if unspecified),
and adds up the results.  The rule's regex is \verb!^([A-Z]([a-z])?\d*)+$!,
which matches:

@@

Parsing a polymer is more difficult because of the repeated monomers: each
monomer's weight needs to be calculated, multiplied by the number of times
it is present in the compound, and that figure added to the sum.  XXX CAN
POLYMERS BE NESTED\@?  Without cascaded parsing the action would need to
maintain state while recursively parsing the monomers embedded in the
polymer.  With cascaded parsing available three rules and actions could be
used as shown in \tableref{Regexes and Actions to calculate the weight of a
chemical compound}:

@@

\begin{table}[ht]
    \caption{Regexes and Actions to calculate the weight of a chemical
    compound}
    \empty{}\label{Regexes and Actions to calculate the weight of a
    chemical compound}
    \begin{tabular}{llp{\chemicalactionwidth}}

        \tabletopline{}%
        No.   & Regex                       & Action \\
        \tablemiddleline{}%
        1     & \verb'((?!H\d+([()]|$))'    & Parse a monomer as described above,  \\
              & \verb' ([A-Z]([a-z])?\d*)+)'& returning \verb!Hnnn!, where \verb!nnn! is
                                              the weight of the monomer.  The first part
                                              of the regex prevents it from parsing the
                                              return value of this action (\verb!Hnnn!),
                                              so the parsing process does not get stuck in
                                              a loop.  \\
        2     & \verb!\(H(\d+)\)(\d+)!      & Parse a repeated monomer that has been
                                              rewritten by the first rule and action,
                                              returning \verb!Hnnn!, where \verb!nnn! is
                                              the product of the first and second numbers
                                              captured by the regex.  \\
        3     & \verb!^H(\d+)+$!            & Repeated applications of rules and
                                              actions one and two will have rewritten the
                                              input to \verb!Hnnn!, where \verb!nnn! is
                                              the weight of the compound.  This action
                                              should do whatever is required with the
                                              calculated weight.  \\
        \tablebottomline{}%

@@

XXX CHECK WITH CARL WHETHER I SHOULD SAY CFL OR CFG IN THE NEXT
PARAGRAPH\@.  XXX DO I NEED TO MENTION CFG/CFL AT ALL\@?

Decoupling the rules from the actions allows external tools to be written
to detect overlapping rules.  Traditional regexes are equivalent in
computational power to \acronym{FA} and can be converted to \acronym{FA},
so regex overlap can be detected by finding a non-empty intersection of two
\acronym{FA}\@.  The standard equation for \acronym{FA} intersection (given
for example in~\cite{intersection-of-NFA-using-Z}) uses DeMorgan's law:
$FA1~\cap{}~FA2~=~\overline{(\overline{FA1}~\cup{}~\overline{FA2})}$, which
has considerable computation complexity.  XXX COMMENT ON THE COMPUTATIONAL
COMPLEXITY OF INVERSION AND UNION\@; ADD A REFERENCE TO THE SPINNING WHEEL
BOOK\@.  Perl 5.10 regexes~\cite{perlre} are more powerful than traditional
regexes: it is possible to match correctly balanced brackets nested to an
arbitrary depth, e.g.\ \verb!/^[^<>]*(<(?:(?>[^<>]+)|(?1))*>)[^<>]*$/!
matches \verb!z<123<pq<>rs>j<r>ml>s!.  Matching balanced brackets requires
the regex engine to maintain state on a stack, so Perl 5.10 regexes are
equivalent in computational power to \acronym{PDA} or a \acronym{CFL};
detecting overlap may require calculating the intersection of two
\acronym{PDA} or \acronyms{CFL} instead of two \acronym{FA}.  The
intersection of two \acronyms{CFL} is not closed, i.e.\ the resulting
language cannot always be parsed by a \acronym{CFL}, so intersection may be
intractable in some cases, e.g.:
$a^{*}b^{n}c^{n}~\cap~a^{n}b^{n}c^{*}~\rightarrow~a^{n}b^{n}c^{n}$.
Detecting overlap amongst $n$ regexes requires calculating $n(n-1)/2$
intersections, resulting in $O(n^2x)$ complexity, where $O(x)$ is the
complexity of calculating \acronym{FA} or \acronym{PDA} intersection.  This
is certainly not a task to be performed every time the parser runs: naive
detection of overlap amongst the Postfix log parser's \numberOFrules{}
rules would require calculating \numberOFruleINTERSECTIONS{} intersections.
When detecting overlap any conditions attached to rules must be taken into
account, because two rules whose regexes overlap may have conditions
attached which prevent the overlap in practice.  A less naive approach to
overlap detection would first check for overlapping conditions amongst
rules, and then check for overlapping regexes amongst the rules in each set
of rules with overlapping conditions.  Depending on the conditions employed
a rule may fall into more than one group, e.g.\ given these three
conditions:

@@

It is possible to define pathological regexes which fall into two main
categories: regexes that match inputs they should not, and regexes that
consume excessive amounts of CPU time during matching.  Defining a regex
that matches inputs it should not is trivial: \verb!/^/! matches the start
of every input.  This regex would be highlighted by a tool that detects
overlapping rules, and would easily be noticed by visual inspection, but
more complex regexes would be harder to spot.  Usually excessive CPU time
is consumed when a regex with a lot of alteration and variable quantifiers
fails to match, but successful matching is generally quite fast
(see~\cite{mastering-regular-expressions} for in-depth discussion).  XXX
THIS NEEDS TO BE EXPANDED, AND GIVE AN EXAMPLE\@.

@@

$\text{\textless{}input\textgreater{}} \mapsto \text{rule-1} |
\text{rule-2} | \text{rule-3} | \dots | \text{rule-n}$

>>>>

XXX GET GREG TO PROOFREAD THIS SECTION\@.

A more complicated example is determining the atomic weight of a compound
given its chemical formula.  A chemical formula is a sequence of one or two
character chemical element abbreviations, each optionally followed by a
number indicating how many of that element are present in the formula (if
omitted the number is assumed to be one).  Some parts of a formula may be
repeated: they are enclosed in parentheses, and the closing parenthesis is
followed by a number indicating how many are present in the formula.  A
monomer is a simple compound, i.e.\ it does not have any repeated portions;
compounds with repeated portions are called polymers.  E.g.\ glucose (a
monomer) has the chemical formula $C_{6}H_{12}O_{6}$ and hydrocarbon (a
polymer) is $CH_{3}{(CH_{2})}_{50}CH_{3}$.  The atomic weight of an
individual element is given in periodic tables, but one element in
particular will be important in the later explanation: Hydrogen,
abbreviated H, has an atomic weight of 1.  The atomic weight of a given
compound is the sum of the atomic weights of each of the elements
multiplied by the number of times they occur; this is simple to calculate
for a monomer, but it is harder for a polymer because the atomic weight of
the repeated monomer must be multiplied by the number of times the monomer
is present in the compound (e.g.\ for hydrocarbon the atomic weight of
$CH_{2}$ must be multiplied by $50$).

Calculating the atomic weight of a monomer requires a single rule and
action: the action separates each element in the monomer, looks up its
atomic weight, multiplies it by the number of times it is present (one if
unspecified), and adds up the results.  The rule's regex is
\verb!^([A-Z]([a-z])?\d*)+$!, which matches:

@@

Parsing a polymer is more difficult because of the repeated monomers: each
monomer's atomic weight needs to be calculated, multiplied by the number of
times it is present in the compound, and that figure added to the sum.
Without cascaded parsing the action would need to maintain state while
recursively parsing the monomers embedded in the polymer.  With cascaded
parsing available three rules and actions could be used as shown in
\tableref{Regexes and Actions to calculate the atomic weight of a chemical
compound}:

@@

\begin{table}[ht]
    \caption{Regexes and Actions to calculate the atomic weight of a
    chemical compound}
    \empty{}\label{Regexes and Actions to calculate the atomic weight of a
    chemical compound}
    \begin{tabular}{llp{\chemicalactionwidth}}

        \tabletopline{}%
        No.   & Regex                       & Action \\
        \tablemiddleline{}%
        1     & \verb'((?!H\d+([()]|$))'    & Parse a monomer as described above,  \\
              & \verb' ([A-Z]([a-z])?\d*)+)'& returning \verb!Hnnn!, where \verb!nnn! is
                                              the atomic weight of the monomer.  The first
                                              part of the regex prevents it from parsing
                                              the return value of this action
                                              (\verb!Hnnn!), ensuring the parsing process
                                              does not get stuck in a loop.  \\
        2     & \verb!\(H(\d+)\)(\d+)!      & Parse a repeated monomer that has been
                                              rewritten by the first rule and action,
                                              returning \verb!Hnnn!, where \verb!nnn! is
                                              the product of the first and second numbers
                                              captured by the regex.  \\
        3     & \verb!^H(\d+)+$!            & Repeated applications of rules and
                                              actions one and two will have rewritten the
                                              input to \verb!Hnnn!, where \verb!nnn! is
                                              the atomic weight of the compound.  This
                                              action should do whatever is required with
                                              the calculated atomic weight.  \\
        \tablebottomline{}%

@@

Decoupling the rules from the actions allows external tools to be written
to detect overlapping rules.  Traditional regexes are equivalent in
computational power to \acronym{FA} and can be converted to \acronym{FA},
so regex overlap can be detected by finding a non-empty intersection of two
\acronym{FA}\@.  The standard equation for \acronym{FA} intersection (given
for example in~\cite{intersection-of-NFA-using-Z}) uses DeMorgan's law:
$FA1~\cap{}~FA2~=~\overline{(\overline{FA1}~\cup{}~\overline{FA2})}$, which
has considerable computation complexity.  XXX COMMENT ON THE COMPUTATIONAL
COMPLEXITY OF INVERSION AND UNION\@; ADD A REFERENCE TO THE SPINNING WHEEL
BOOK\@.  Perl 5.10 regexes~\cite{perlre} are more powerful than traditional
regexes: it is possible to match correctly balanced brackets nested to an
arbitrary depth, e.g.\ \verb!/^[^<>]*(<(?:(?>[^<>]+)|(?1))*>)[^<>]*$/!
matches \verb!z<123<pq<>rs>j<r>ml>s!.  Matching balanced brackets requires
the regex engine to maintain state on a stack, so Perl 5.10 regexes are
equivalent in computational power to \acronym{PDA}; detecting overlap may
require calculating the intersection of two \acronym{PDA} instead of two
\acronym{FA}.  The intersection of two \acronym{PDA} is not closed, i.e.\
the result cannot always be implemented using a \acronym{PDA}, so
intersection may be intractable in some cases, e.g.:
$a^{*}b^{n}c^{n}~\cap~a^{n}b^{n}c^{*}~\rightarrow~a^{n}b^{n}c^{n}$.
Detecting overlap amongst $n$ regexes requires calculating $n(n-1)/2$
intersections, resulting in $O(n^2x)$ complexity, where $O(x)$ is the
complexity of calculating \acronym{FA} or \acronym{PDA} intersection.  This
is certainly not a task to be performed every time the parser runs: naive
detection of overlap amongst \parsernames{} \numberOFrules{} rules would
require calculating \numberOFruleINTERSECTIONS{} intersections.  When
detecting overlap any conditions attached to rules must be taken into
account, because two rules whose regexes overlap may have conditions
attached which prevent the overlap in practice.  A less naive approach to
overlap detection would first check for overlapping conditions amongst
rules, and then check for overlapping regexes amongst the rules in each set
of rules with overlapping conditions.  Depending on the conditions employed
a rule may fall into more than one group, e.g.\ given these three
conditions:

@@

It is possible to define pathological regexes which fall into two main
categories: regexes that match inputs they should not, and regexes that
consume excessive amounts of CPU time during matching.  Defining a regex
that matches inputs it should not is trivial: \verb!/^/! matches the start
of every input.  This regex would be found by a tool that detects
overlapping rules, and would easily be noticed by visual inspection, but
more complex regexes would be harder to spot.  Regexes that match more than
they should are a problem not because of excessive resource usage, but
because they may prevent the correct rule from recognising the input.  If
an adaptive ordering system is used (see \sectionref{rule ordering for
efficiency}) then a rule with a regex that matches too much may be promoted
up through the list, displacing an increasing number of correct rules as it
moves.

Usually excessive CPU time is consumed when a regex with a lot of
backtracking, caused by alteration and/or nested quantifiers, fails to
match; successful matching is generally quite fast with such regexes, so
problematic regexes may go unnoticed for some time.  E.g.\ with most regex
engines matching double quoted strings with \verb!"([^"\\]+|\\.)*"! is very
fast when there is a match, but when the match fails its computational
complexity is $O(2^{n})$; see~\cite{mastering-regular-expressions} for
in-depth discussion of nested quantifiers, backtracking, alteration, and
capturing groups.  Pathological regexes which consume excessive CPU time
can be difficult to detect, both by visual inspection and by machine
inspection, but if a regex is converted to a \acronym{FA} or the internal
representation used by the regex engine, it may be possible to determine if
there are nested quantifiers or other troublesome constructs present.
Modern regex engines have addressed many of these problems, e.g.\ the regex
to match double quoted strings given above fails immediately with Perl
5.10, regardless of the input length, because the regex engine looks for
the required double quotes first.  Similarly Perl 5.10's regex engine
optimises alterations starting with literal text into a trie, which has
constant matching time, whereas the matching time for alteration is
proportional to the number of alternatives.  Perl 5.10 regexes can use
\verb!(?>pattern)!, which matches \verb!pattern! the first time the regex
engine passes over it, and does not change what it originally matched if
the regex engine backtracks over it, so troublesome regexes can be modified
to use this to reduce the impact of backtracking; Prolog
(\urlLastCheckedNoParens{http://en.wikipedia.org/wiki/Prolog}{2009/03/03})
users will notice a similarity to the \verb'!' (cut) operator.  A
presentation showing some of Perl 5.10's regex features is available at
\urlLastChecked{http://www.regex-engineer.org/slides/perl510_regex.html}{2009/03/03}.

@@

$\text{\textless{}input\textgreater{}} \mapsto{} \text{rule-1} |
\text{rule-2} | \text{rule-3} | \dots | \text{rule-n}$
